Gut microbes and their genes are associated with cognition and neuroanatomy in children

### 2023-03-15 manuscript/main.tex
through figure 2

To begin to address this need, we investigated trends and associations between
the gut microbiome and
neurocognitive development in a large cohort of healthy and neurotypically-developing
children from infancy through 10 years of age. Gut microbial
communities were assessed using shotgun metagenomic sequencing, enabling
profiling at both the taxonomic and gene-functional level.
Cognitive skills and abilities were measured using age-appropriate
psychometric assessments of cognitive function s, i.e.,
the Mullen Scales of Early Learning (MSEL) \cite{mullenMullenScalesEarly1995},
Weschler Preschool and Primary Scale of Intelligence, 4th Edition (WPPSI-4) 
\cite{wechslerWechslerPreschoolPrimary2012}
and Weschler Intelligence Scale for Children, 5th Edition (WISC-V)
\cite{wechslerWechslerIntelligenceScale1949}.
Finally, we also assessed emerging brain structure magnetic resonance imaging
(MRI). Through a combination of classical statistical analysis and machine
learning, we find that the development of the gut microbiome,
children's cognitive abilities, and brain structure are intimately linked, with
both microbial taxa and gene functions able to predict cognitive
performance and brain structure.

@@

We investigated the co-development of the brain and the microbiome in
361 healthy and neurotypically-developing children (163 female) between 82 days and 10 years of age
(Table 1,  Figure 1A, Supplementary Figure 1)\cite{forrestAdvancingScienceChildren2018}
using a variety of orthogonal microbial and neurocognitive measures.
These included shotgun metagenomic sequencing,
age-appropriate cognitive and behavioral assessments
using full-scale composite scores from the MSEL, WPPSI-4 and WISC-V,
and neuroimaging measures of cortical and sub-cortical morphology. 
Different assessment instruments were used depending on the child's age
due to the unique age-range and psychometric properties measured by each
(i.e., the MSEL for children under 4 years of age, The WPPSI-3 for 3-6 year-olds,
and the WISC-V for children older than 6), but were normalized to a common scale (Figure 1B),
As expected, the greatest differences in microbial taxa were observed across
child age (Figure 1C-D, PCoA axis 1), while older children were primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-\emph{Prevotella copri} (Supplementary Figure 2). Overall variation
in gut microbial genes and also in brain volume profiles was similarly
driven largely by subject age (Figure 1E-F).

@@

\begin{table}[!htb]
    \centering
    \begin{tabular}{|r|r|r|r|r|}
      \hline\hline
      \textbf{group} & \textbf{subgroup} & \textbf{all} & \textbf{under6mo} & \textbf{over18mo} \\\hline
      N subjects &  & 361 & 74 & 257 \\  \hline
    \multirow{3}{*}{Samples} & 1 & 262 (72.6\%) & 74 (100.0\%) & 203 (79.0\%) \\ \cline{2-5}
         & 2 & 72 (19.9\%) & 0 (0.0\%) & 46 (17.9\%) \\ \cline{2-5}
         & \textgreater 2 & 27 (7.5\%) & 0 (0.0\%) & 8 (3.1\%) \\ \hline
    \multirow{3}{*}{Age (months)} & min & 2.76 & 2.76 & 18.03 \\ \cline{2-5}
         & max & 119.3 & 6.0 & 119.3 \\ \cline{2-5}
         & median & 25.79 & 3.66 & 45.92 \\ \hline
    \multirow{2}{*}{Sex} & F & 163 (45.2\%) & 37 (50.0\%) & 114 (44.4\%) \\   \cline{2-5}
                         & M & 198 (54.8\%) & 37 (50.0\%) & 143 (55.6\%) \\  \hline
    \multirow{6}{*}{Race} & White & 252 (69.8\%) & 42 (56.8\%) & 191 (74.3\%) \\   \cline{2-5}
        & Black & 40 (11.1\%) & 16 (21.6\%) & 21 (8.2\%) \\ \cline{2-5}
        & Indiginous & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) \\ \cline{2-5}
        & Asian & 7 (1.9\%) & 1 (1.4\%) & 6 (2.3\%) \\ \cline{2-5}
        & Mixed & 48 (13.3\%) & 12 (16.2\%) & 29 (11.3\%) \\ \cline{2-5}
        & Other & 2 (0.6\%) & 1 (1.4\%) & 2 (0.8\%) \\ \hline
    \multirow{6}{*}{Maternal Ed.} & Junior high school & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) \\   \cline{2-5}
        & Some high school & 6 (1.7\%) & 3 (4.1\%) & 4 (1.6\%) \\ \cline{2-5}
        & High school grad & 37 (10.2\%) & 13 (17.6\%) & 16 (6.2\%) \\ \cline{2-5}
        & Some college & 95 (26.3\%) & 25 (33.8\%) & 57 (22.2\%) \\ \cline{2-5}
        & College grad & 90 (24.9\%) & 17 (23.0\%) & 69 (26.8\%) \\ \cline{2-5}
        & Grad/professional school & 123 (34.1\%) & 15 (20.3\%) & 104 (40.5\%) \\\hline\hline
    \end{tabular}
    \caption{\label{tab:demo}Subject demographics}
\end{table}

As several prior studies have demonstrated links between specific gut taxa
with ASD \cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021},
adult cognitive skill (e.g., executive functioning)\cite{magnussonRelationshipsDietrelatedChanges2015},
and adult mental health (both anxiety and depressive symptioms)\cite{mayneris-perxachsMicrobiotaAlterationsProline2022,needhamGutderivedMetaboliteAlters2022}
we sought to determine if specific taxa or gene functions were linked with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism are associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age) separately.
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was associated with a small but significant variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0124, q \textless{} 0.001),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0119, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants.
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0215, and  gene functions
annotated with UniRef90s - clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0245, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging profiles (R\textsuperscript{2} = 0.256, q
\textless{} 0.001).

@@

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive ability score (Table 2, Figure 2C-G) and found that
several metabolic pathways were significantly enriched or depleted in
children with higher cognitive ability scores. This was true both when
considering all ages together, though the enrichment of most
pathways was more pronounced either in children under 6 months or those over 18
months. For example, genes for degrading the 3-carbon SCFA propionate
were significantly depleted in children with higher cognitive ability
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
In children over 18 months old, synthesis of menaquinone (vitamin K)
was also negatively associated with cognitive function score (menaquinone synthesis I,
E.S. = -0.170, q = 0.0183), while
genes for the synthesis of the amino acids
glutamate and tryptophan were significantly enriched in children with
higher cognitive function scores (Glutamate synthesis I, E.S. = 0.242, 
q = 0.047; Tryptophan synthesis, E.S. = 0.119, q = 0.041).
Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.

>>>>

To begin to address this need, we investigated trends and associations between
the gut microbiome and
neurocognitive development in a large cohort of healthy and neurotypically-developing
children from infancy through 10 years of age. Gut microbial
communities were assessed using shotgun metagenomic sequencing, enabling
profiling at both the taxonomic and gene-functional level.
Cognitive skills and abilities were measured using age-appropriate
psychometric assessments of cognitive function s, i.e.,
the Mullen Scales of Early Learning (MSEL) \cite{mullenMullenScalesEarly1995},
Weschler Preschool and Primary Scale of Intelligence, 4th Edition (WPPSI-4) 
\cite{wechslerWechslerPreschoolPrimary2012}
and Weschler Intelligence Scale for Children, 5th Edition (WISC-V)
\cite{wechslerWechslerIntelligenceScale1949}.
Finally, we also assessed emerging brain structure using magnetic resonance imaging
(MRI). Through a combination of classical statistical analysis and machine
learning, we find that the development of the gut microbiome,
children's cognitive abilities, and brain structure are intimately linked, with
both microbial taxa and gene functions able to predict cognitive
performance and brain structure.

@@

We investigated the co-development of the brain and the microbiome in
361 healthy and neurotypically-developing children (163 female) between 82 days and 10 years of age
(Table 1,  Figure 1A, Supplementary Figure 1)
using a variety of orthogonal microbial and neurocognitive measures.
These included shotgun metagenomic sequencing,
age-appropriate cognitive and behavioral assessments
using full-scale composite scores from the MSEL, WPPSI-4 and WISC-V,
and neuroimaging measures of cortical and sub-cortical morphology. 
Different assessment instruments were used depending on the child's age
due to the unique age-range and psychometric properties measured by each
(i.e., the MSEL for children under 4 years of age, The WPPSI-3 for 3-6 year-olds,
and the WISC-V for children older than 6), but were normalized to a common scale (Figure 1B),
As expected, the greatest differences in microbial taxa were observed across
child age (Figure 1C-D, PCoA axis 1), while older children were primarily
stratified into Bacteroidetes-dominant, Firmicutes-dominant, or
high-abundance \emph{Prevotella copri} (Supplementary Figure 2).
Overall subject-to-subject variation in gut microbial genes
and brain volume profiles was similarly driven largely by subject age (Figure 1E-F).

@@

\begin{table}[!htb]
    \centering
    \begin{tabular}{|r|r|r|r|r|}
      \hline\hline
      \textbf{group} & \textbf{subgroup} & \textbf{all} & \textbf{under6mo} & \textbf{over18mo} \\\hline
      N subjects &  & 361 & 74 & 257 \\  \hline
    \multirow{3}{*}{Samples} & 1 & 262 (72.6\%) & 74 (100.0\%) & 203 (79.0\%) \\ \cline{2-5}
         & 2 & 72 (19.9\%) & 0 (0.0\%) & 46 (17.9\%) \\ \cline{2-5}
         & \textgreater 2 & 27 (7.5\%) & 0 (0.0\%) & 8 (3.1\%) \\ \hline
    \multirow{3}{*}{Age (months)} & min & 2.76 & 2.76 & 18.03 \\ \cline{2-5}
         & max & 119.3 & 6.0 & 119.3 \\ \cline{2-5}
         & median & 25.79 & 3.66 & 45.92 \\ \hline
    \multirow{2}{*}{Sex} & F & 163 (45.2\%) & 37 (50.0\%) & 114 (44.4\%) \\   \cline{2-5}
                         & M & 198 (54.8\%) & 37 (50.0\%) & 143 (55.6\%) \\  \hline
    \multirow{6}{*}{Race} & White & 252 (69.8\%) & 42 (56.8\%) & 191 (74.3\%) \\   \cline{2-5}
        & Black & 40 (11.1\%) & 16 (21.6\%) & 21 (8.2\%) \\ \cline{2-5}
        & Indiginous & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) \\ \cline{2-5}
        & Asian & 7 (1.9\%) & 1 (1.4\%) & 6 (2.3\%) \\ \cline{2-5}
        & Mixed & 48 (13.3\%) & 12 (16.2\%) & 29 (11.3\%) \\ \cline{2-5}
        & Other & 2 (0.6\%) & 1 (1.4\%) & 2 (0.8\%) \\ \hline
    \multirow{6}{*}{Maternal Ed.} & Junior high school & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) \\   \cline{2-5}
        & Some high school & 6 (1.7\%) & 3 (4.1\%) & 4 (1.6\%) \\ \cline{2-5}
        & High school grad & 37 (10.2\%) & 13 (17.6\%) & 16 (6.2\%) \\ \cline{2-5}
        & Some college & 95 (26.3\%) & 25 (33.8\%) & 57 (22.2\%) \\ \cline{2-5}
        & College grad & 90 (24.9\%) & 17 (23.0\%) & 69 (26.8\%) \\ \cline{2-5}
        & Grad/professional school & 123 (34.1\%) & 15 (20.3\%) & 104 (40.5\%) \\\hline\hline
    \end{tabular}
    \caption{\label{tab:demo}Subject demographics of the ECHO RESONANCE cohort.}
\end{table}

As several prior studies have demonstrated links between specific gut taxa
with ASD \cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021},
adult cognitive skill (e.g., executive functioning)
\cite{magnussonRelationshipsDietrelatedChanges2015},
and adult mental health (both anxiety and depressive symptioms)
\cite{mayneris-perxachsMicrobiotaAlterationsProline2022,needhamGutderivedMetaboliteAlters2022},
we sought to determine if specific taxa or gene functions were linked
with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism were associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age) separately.
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was associated with a small but significant variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0124, q \textless{} 0.001),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0119, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants (Figure 1G).
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0215, and  gene functions
annotated with UniRef90s - clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0245, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging profiles (R\textsuperscript{2} = 0.256, q
\textless{} 0.001).

@@

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive ability score (Table 2, Figure 2C-G) and found that
several metabolic pathways were either significantly enriched or depleted in
children with higher cognitive ability scores. This was true both when
considering all ages (0-120 months) together, though the enrichment of most
pathways was more pronounced either in children under 6 months or those over 18
months. For example, genes for degrading the 3-carbon SCFA, propionate,
were significantly depleted in children with higher cognitive ability
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
In children over 18 months old, synthesis of menaquinone (vitamin K)
was also negatively associated with cognitive function score (menaquinone synthesis I,
E.S. = -0.170, q = 0.0183), while
genes for the synthesis of the amino acids
glutamate and tryptophan were significantly enriched in children with
higher cognitive function scores (Glutamate synthesis I, E.S. = 0.242, 
q = 0.047; Tryptophan synthesis, E.S. = 0.119, q = 0.041).
Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.
